\textit{MSF: Modulated Sub-graph Finder}

### 2018-04-29 MSF_Paper_FA.tex
Comparison correction with figure

On these grounds we propose a novel approach to make use of the rich
gene and protein interaction annotation resources available to gain
additional functional insights from basic DGEA. To this, we start with
the presupposition that expression of neighboring genes within a
functional pathway are not independent from each other. Rather, they
are often regulating each others expression or are part of the same
regulon~\cite{Michalak}. We understand that the categorization of
links between genes into labeled pathways is often an arbitrary one,
given the extensive cross talk between different pathways. Although
these categories have proven to be useful in many situations, they
force a certain perspective onto the interpretation of novel
data. Based on these two principles, we aim to find sub-graphs of
connected genes within cell signaling network which exhibit as a whole
significant differential expression changes. This approach differes in
two main aspects from common pathway analysis tools. First, it does
not aim to identify functional pathways enriched in differentially
expressed genes, but detects sub-graphs or branches in a network graph
(potentially spanning more than one functionally grouped pathway)
which is coherently modulated. Second, it aims to improve the DGEA on
the gene level, by collecting the information of neighboring genes,
which as a whole might exhibit prominent enough signal to be called,
again as a whole, significantly modulated.

@@

All of this can be helpful to understand the cause and effect of a
stimulus and might inform about potential points of intervention. The
proposed algorithm was implemented as a java program, which was named
\texttt{Modulated Sub-graph Finder} (abbreviated
\texttt{MSF}). \texttt{MSF} can help transform the information
obtained from DGEA into comprehensible knowledge of signal
transduction of genes andl thereby being a valuable complement to
existing pathway based methods. \texttt{MSF} is freely accessible on
github under the terms of the Creative Commons Attribution 4.0
International License.

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to find
concertedly modulated sub-graphs in networks of biological
interactions.  \texttt{MSF} does not use predefined gene sets grouped
into functional units, but rather relies purely on the network of
interacting genes.  The inputed network consists of nodes
corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

@@

EBOV infection count data was downloaded from GEO (GSE84188)
\TODO{citation needed};\TODO{What was infected?}. Differential gene
expression analysis was performed on the count data with edgeR package
(version 3.4.2)~\cite{edgeR}. The DEG analysis results generated by
edgeR were used as input for \texttt{MSF}. The EBOV infection
experiments describe the course of infection at three time-points 6,
24 and 48 hour post infection (hpi). For the earliest time point at
6~hpi, five modulated sub-graphs were identified with 41, 107, and 18
genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10,
CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as a subset
of CD4+ T helper cells. IFNB1 and IFNA1 were both identified as two of
the possible sources. Most of the sources identified by \texttt{MSF}
were type~I interferon induced genes. At 24~hpi nine modulated
sub-graphs were identified with four main sub-graphs consisting of 38,
85, 148 and 167 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point
48~hpi, eleven modulated sub-graphs were identified. Eight of the
sub-graphs were less than seven genes and main sub-graphs had 39, 71
and 210 genes. IFNB1 and IFNA1 were identified as the two sources out
of the possible sources. Table 1 shows if \texttt{MSF} does not
consider the complete DEG results and uses different cut offs for
significant genes from DEG results, how the signal transduction would
fall apart into meaningless small sub-graphs and single unconnected
genes.

@@

\TODO{I hoped that you would not just add the table as discussed but
  would also add some discussion in the main text explaining what the
  table~\ref{tab:rawVsHartung} illustrates!}

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6~hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. The gene based graphs on
 the right hand side, represent the immediate output of the
 \texttt{MSF}-analysis, visualised by StringApp~\cite{StringApp} in
 Cytoscape~\cite{Cyto}. Each node represents a gene part of a
 modualted sub-graph, whereby the associated colors code the
 functional annotation deduced from KEGG Pathways. The crosstalk
 between the pathways and also the mutliple employment of many genes
 is evident. The more schematic drawing on the right side represents
 the effortlessly deduced flow of information between the sensors and
 effectors in this particular example.

 In more detail, sub-graph~1 (top) shows how the activation of
 toll-like receptor, cytokine, chemokine and jak-stat genes lead via
 TNF into apoptosis. The next significant sub-grap (sub-graph~2:
 middle) reveals how information from the Extra-cellular matrix (ECM)
 receptor, which are reported to interact with Ebola glycoprotein
 (GP)~\cite{Veljkovic}, chemokines and cytokines, and cytosolic DNA
 sensing, is integrated into again modulation of apoptosis
 pathway. Eventually, sub-graph~3 (bottom) demonstrate how INFA1 and
 INFB1 modulate once more, via only a few intermediat steps, the
 aptoptotic response of the cell.

@@

 To assess the robustness and stability of our method, Poisson
 distributed noise was added to the read counts of the three
 time-points data set, used in the previous analysis. Then DGEA was
 carried out on the disturbed data with the same parameters as for the
 native data using edgeR, followed by analysis with \texttt{MSF}. This
 procedure was carried out 100 times.  Every time the genes from the
 modulated sub-graphs identified from noisy data were compared to the
 genes of sub-graphs identified from the native data. This was also
 done for the DEG obtained for each run, a cutoff of FDR less than
 0.05 was used to compare the retrieved DEG from noisy data to the
 native data.  The robustness of \texttt{MSF} and the DEG analysis for
 the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can
 be considered as rather stringent, which is already reflected limited
 recall rate in the edgeR based DGEA, between \TODO{X and Y~\%}. For
 \texttt{MSF}-analysis the observed median recall rates lay between
 71~\% (6~hpi) and 84~\% (48~hpi).

\subsection*{Comparison to existing pathway analysis tools}

 A comparison to other tools was performed on the same dataset using
 each software's default parameters. We compared the gene enrichment
 analysis \TODO{does MSF realy do a ``gene enrichment analysis''?} in
 \texttt{MSF} identified sub-graphs to pathway analysis by
 SPIA~\cite{Tarca} for KEGG and Reactome pathway enrichment analysis
 for Reactome.  We used Cytoscape~\cite{Cyto} plugins to identify the
 enriched patwhays, CytoKegg~\cite{Cytokegg} for KEGG and Reactome
 FI~\cite{Reactome} for Reactome for \texttt{MSF} identified
 sub-graphs. \TODO{I do not get those two sentences. What do those
   tools do? How do you compare the results of those tools?}

At 6 hpi Cytokegg identified the sub-graph genes to be enriched in
Toll-like receptor signaling, TNF signaling, IL-17 signaling and
NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The
earliest response EBOV induces on cytokine is via TLR4-mediated
signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes
at 6 hpi showed toll-like receptor signaling being most significantly
dis-regulated pathway, this important signaling pathway was not
identified by SPIA along with RIG-I-like receptor signaling pathway,
TNF signaling pathway and IL-17 signaling. On the later time-points
\texttt{MSF} and SPIA showed consensus on the important pathways e.g
Chemokine signaling, Cytokine-Cytokine receptor interaction, NF-kappa
B signaling, Cytosolic DNA-sensing and RIG-I-like receptor signaling
as being dis-regulated.

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that \texttt{MSF} is able to identify the
important modulated genes of the early infection stage. \TODO{The
  whole paragraph is hard to read and get the essentials. I suggest
  you describe in this section what the different tools work,
  especially what are the differences between them. For the results I
  suggest a plot; maybe something in the line of upset or a venn
  diagram, if appropriate.}

@@

Classic pathway analysis tools aim to detect in lists of significantly
deregulated genes enriched associations with pathway genes categorized
by their biological function and their interactions. Thereby,
depending on the tool, the internal pathway topology is considered or
neglected alltogether. The here presented tool, \texttt{MSF}, employs
a different approach, by aiming to detect sub-graphs in whole gene
regulatory networks which are significantly deregulated in a concerted
manner. To this end, neighboring genes in the user provided network
are tested for jointly common regulation. Exploiting that each gene's
abundance, although not independent from its neighbors, is measured
repeatedly on its own, sensitivity can be increased by our applied
p-value meta-analyse, namely Hartung's method. This potentially
enables to call just not significant modulated genes based on the DGEA
to be convincingly called to be part of a deregulated gene group.
Furthermore, it allows to identify connected sub-graphs, representing
the propagation of gene regulation perturbation in the input
network. A better understanding of this propagation, especially the
critical spots such as sensors, effectors, or intermediate bottlenecks
and hubs, facilitates the projection of potential intervention points,
e.g. for drug development. Since \texttt{MSF} only uses interaction
information in gene regulation network, but not the functional
grouping of the genes into functional pathways, it is especially
adapted to discover so called cross-talk between such pathways.

@@

\texttt{MSF} is a fast, robust and easy to use tool to find
concertedly modulated sub-graphs in a given network. Its
implementation in java enables it use across many operating
systems. It requires as input the results from a differential gene
expression analysis in the appropriate file format. So far the raw
output from edgeR and DESeq2~\cite{love2014moderated} are supported
\TODO{did you already implement a input option for
  DESeq2?}. Furthermore, a gene network has to be provided \TODO{do
  you check at some point if the IDs between the DGEA input and the
  gene network fit?} in the format of a simple adjacency list \TODO{is
  this true?}. For convenience \texttt{MSF} is distributed with a
prefiltered version of all directed cell signalling interactions from
Reactome Functional interactions (FIs) Version 2016. \TODO{citation
  needed} ;\TODO{did you check if reactome comes with the appropriate
  licence that we allowed to redistribute this file? this is very
  important otherwise we need good laywers} ;\TODO{In the case study
  section you never mention this file and especially the prefiltering
  of it? wouldn t you consider this crucial?}

@@

The Ebola infection RNAseq dataset analyzed during the current study
are available in the GEO repository (GSE84188) \TODO{citation
  needed}. The cell signaling network file used is from Reactome
Functional interactions (FIs) Version 2016 \TODO{citation needed}.

\subsection*{Data and Software Availability}

\TODO{see
  \url{https://f1000research.com/for-authors/article-guidelines/software-tool-articles}
  what has to be stated there}

\section*{Author contributions}
Conceptualization: FA, ILH; Funding Acquisition: ILH; Resources: ILH;
Software: MRF; Writing – Original Draft Preparation: MRF; Writing –
Review \& Editing: FA, ILH. All authors read and approved the final
manuscript.

@@

\subsection*{Acknowledgements}

We thank all our colleagues who provided insight and expertise that
greatly assisted the research and all reviewers for their constructive
criticism.\TODO{I'd suggest name those colleagues by name or remove
  the section.}

@@

>>>>

On these grounds we propose a novel approach to make use of the rich
gene and protein interaction annotation resources available to gain
additional functional insights from basic DGEA. To this, we start with
the presupposition that expression of neighboring genes within a
functional pathway are not independent from each other. Rather, they
are often regulating each others expression or are part of the same
regulon~\cite{Michalak}. We understand that the categorization of
links between genes into labeled pathways is often an arbitrary one,
given the extensive cross talk between different pathways. Although
these categories have proven to be useful in many situations, they
force a certain perspective onto the interpretation of novel
data. Based on these two principles, we aim to find sub-graphs of
connected genes within cell signaling network which exhibit as a whole
significant differential expression changes. This approach differs in
two main aspects from common pathway analysis tools. First, it does
not aim to identify functional pathways enriched in differentially
expressed genes, but detects sub-graphs or branches in a network graph
(potentially spanning more than one functionally grouped pathway)
which is coherently modulated. Second, it aims to improve the DGEA on
the gene level, by collecting the information of neighboring genes,
which as a whole might exhibit prominent enough signal to be called,
again as a whole, significantly modulated.

@@

All of this can be helpful to understand the cause and effect of a
stimulus and might inform about potential points of intervention. The
proposed algorithm was implemented as a java program, which was named
\texttt{Modulated Sub-graph Finder} (abbreviated
\texttt{MSF}). \texttt{MSF} can help transform the information
obtained from DGEA into comprehensible knowledge of signal
transduction of genes and thereby being a valuable complement to
existing pathway based methods. \texttt{MSF} is freely accessible on
github under the terms of the Creative Commons Attribution 4.0
International License.

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to find
concertedly modulated sub-graphs in networks of biological
interactions.  \texttt{MSF} does not use predefined gene sets grouped
into functional units, but rather relies purely on the network of
interacting genes.  The inputted network consists of nodes
corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

@@

EBOV infection count data was downloaded from GEO (GSE84188). Differential gene
expression analysis was performed on the count data with edgeR package
(version 3.4.2)~\cite{edgeR}. The DEG analysis results generated by
edgeR were used as input for \texttt{MSF}. Directed cell signaling interactions were filtered from
Reactome Functional interactions (FIs) Version 2016~\cite{Cytokegg}, which was used as a second input for \texttt{MSF}.

The EBOV infection
experiments describe the course of infection at three time-points 6,
24 and 48 hour post infection (hpi). For the earliest time point at
6~hpi, five modulated sub-graphs were identified with 41, 107, and 18
genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10,
CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as a subset
of CD4+ T helper cells. IFNB1 and IFNA1 were both identified as two of
the possible sources. Most of the sources identified by \texttt{MSF}
were type~I interferon induced genes. At 24~hpi nine modulated
sub-graphs were identified with four main sub-graphs consisting of 38,
85, 148 and 167 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point
48~hpi, eleven modulated sub-graphs were identified. Eight of the
sub-graphs were less than seven genes and main sub-graphs had 39, 71
and 210 genes. IFNB1 and IFNA1 were identified as the two sources out
of the possible sources.

@@

Table~\ref{tab:rawVsHartung} shows if \texttt{MSF} does not
consider the complete DEG results and uses different cut offs for
significant genes from DEG results, how the signal transduction would
fall apart into incomprehensible small sub-graphs and single unconnected
genes.

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6~hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. The gene based graphs on
 the right hand side, represent the immediate output of the
 \texttt{MSF}-analysis, visualized by StringApp~\cite{StringApp} in
 Cytoscape~\cite{Cyto}. Each node represents a gene part of a
 modulated sub-graph, whereby the associated colors code the
 functional annotation deduced from KEGG Pathways. The cross-talk
 between the pathways and also the multiple employment of many genes
 is evident. The more schematic drawing on the right side represents
 the effortlessly deduced flow of information between the sensors and
 effectors in this particular example.

 In more detail, sub-graph~1 (top) shows how the activation of
 toll-like receptor, cytokine, chemokine and jak-stat genes lead via
 TNF into apoptosis. The next significant sub-graph (sub-graph~2:
 middle) reveals how information from the Extra-cellular matrix (ECM)
 receptor, which are reported to interact with Ebola glycoprotein
 (GP)~\cite{Veljkovic}, chemokines and cytokines, and cytosolic DNA
 sensing, is integrated into again modulation of apoptosis
 pathway. Eventually, sub-graph~3 (bottom) demonstrate how INFA1 and
 INFB1 modulate once more, via only a few intermediate steps, the
 apoptotic response of the cell.

@@

 To assess the robustness and stability of our method, Poisson
 distributed noise was added to the read counts of the three
 time-points data set, used in the previous analysis. Then DGEA was
 carried out on the disturbed data with the same parameters as for the
 native data using edgeR, followed by analysis with \texttt{MSF}. This
 procedure was carried out 100 times.  Every time the genes from the
 modulated sub-graphs identified from noisy data were compared to the
 genes of sub-graphs identified from the native data. This was also
 done for the DEG obtained for each run, a cutoff of FDR less than
 0.05 was used to compare the retrieved DEG from noisy data to the
 native data.  The robustness of \texttt{MSF} and the DEG analysis for
 the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can
 be considered as rather stringent, which is already reflected limited
 recall rate in the edgeR based DGEA, between 84~\% (6~hpi) and 93~\% (24~hpi). For
 \texttt{MSF}-analysis the observed median recall rates lay between
 71~\% (6~hpi) and 84~\% (48~hpi).

\subsection*{Comparison to Reactome pathway analysis tool}

  Gene enrichment analysis was performed on \texttt{MSF} identified sub-graphs for each time-point using Reactome Analyze data tool~\cite{Reactome}. Reactome's over-representation analysis tests whether certain Reactome pathways are enriched by the list of genes submitted. For comparison the DEG results from edgeR were filtered using a cut off of adjusted p-value < 0.1. This subset of DEG list was used for gene enrichment analysis. Enriched pathways were further filtered by confidence score p-values, pathways having p-value less $<$ 0.05 for cut off DEG list and $>$ 1$^{-10}$ for  pathway enrichment in \texttt{MSF} identified sub-graph genes. Out of 282 pathways dis-regulated in the two comparison analysis with three time-points each, pathways that were present in at least three time-points are shown in Fig~\ref{fig:msfvsreactome}.

  The comparison shows most of the important pathways known from literature to be dis-regulated by Ebola infection are enriched in both the enrichment analysis. Toll-Like receptor signaling pathway when interacts with EBOV glycoprotein (GP), triggers the activation of cytokines~\cite{Olejnik}. This pathway was not identified when a p-value cut off DEG list was analyzed for enrichment. Due to \texttt{MSF} considering the complete DEG results, even the weak signal at the earliest time-point was detected; for example Toll-like receptor signaling.

@@

Classic pathway analysis tools aim to detect in lists of significantly
deregulated genes enriched associations with pathway genes categorized
by their biological function and their interactions. Thereby,
depending on the tool, the internal pathway topology is considered or
neglected all together. The here presented tool, \texttt{MSF}, employs
a different approach, by aiming to detect sub-graphs in whole gene
regulatory networks which are significantly deregulated in a concerted
manner. To this end, neighboring genes in the user provided network
are tested for jointly common regulation. Exploiting that each gene's
abundance, although not independent from its neighbors, is measured
repeatedly on its own, sensitivity can be increased by our applied
p-value meta-analysis, namely Hartung's method. This potentially
enables to call just not significant modulated genes based on the DGEA
to be convincingly called to be part of a deregulated gene group.
Furthermore, it allows to identify connected sub-graphs, representing
the propagation of gene regulation perturbation in the input
network. A better understanding of this propagation, especially the
critical spots such as sensors, effectors, or intermediate bottlenecks
and hubs, facilitates the projection of potential intervention points,
e.g. for drug development. Since \texttt{MSF} only uses interaction
information in gene regulation network, but not the functional
grouping of the genes into functional pathways, it is especially
adapted to discover so called cross-talk between such pathways.

@@

\texttt{MSF} is a fast, robust and easy to use tool to find
concertedly modulated sub-graphs in a given network. Its
implementation in java enables it use across many operating
systems. It requires as input the results from a differential gene
expression analysis in the appropriate file format. So far the raw
output from edgeR~\cite{edgeR} and DESeq2~\cite{love2014moderated} are supported
. Furthermore, a gene network has to be provided in the format of a directed adjacency list.

@@

The Ebola infection RNAseq dataset analyzed during the current study
are available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell signaling network file used is from Reactome
Functional interactions (FIs) Version 2016~\cite{Cytokegg}

\subsection*{Software Availability}

Source code: https://github.com/MariamFarman/Modulated-SubGraph-Finder \newline
Software license: Creative Commons Attribution 4.0 International License
\section*{Author contributions}
Conceptualization: FA, ILH; Funding Acquisition: ILH; Resources: ILH;
Software: MRF; Writing – Original Draft Preparation: MRF; Writing –
Review \& Editing: FA, ILH. All authors read and approved the final
manuscript.

@@

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactome}
	\caption{The heatmap shows Reactome pathway enrichment for MSF identifed sub-graphs for the three time-points and the Reactome pathway enrichment for subset of DEG with adjusted p-value cut off of < 0.1. Green color shows the pathway being enriched and red color shows pathway not found as enriched.}
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}
